# Modeling Port Relationships 

```{r, include=FALSE}
library(lattice)
library(reshape)
library(igraph)
library(ggbiplot)
library(gridExtra)
setwd("~/Desktop/Stats Thesis/thesis-sp18-wu-anomalydet/index")
argus = readRDS("data/argus_complete.rds")
means = readRDS("data/means.RDS")
vars = readRDS("data/vars.RDS")
freqs = readRDS("data/freqs.RDS")
combos = readRDS("data/combos.RDS")

n_Sport = 20
n_Dport = 20

#get freqs
Sport_table = as.data.frame(table(argus$Sport))
Sport_table = Sport_table[order(-Sport_table$Freq),]
top_Sport = (head(Sport_table$Var1, n_Sport))

#get freqs
Dport_table = as.data.frame(table(argus$Dport))
Dport_table = Dport_table[order(-Dport_table$Freq),]
top_Dport = (head(Dport_table$Var1, n_Dport))
```

The following properties of $T$ inform the matrix and tensor completion strategies in chapters 3, 4, and 5.

## Missingness
```{r, echo = FALSE, fig.height = 5, fig.width = 5}
### MISSINGNESS VISUALIZATIONS

#sum(apply(means, MARGIN = 1, FUN = function(x) length(x[is.na(x)])))

missing = matrix(list(), nrow = n_Sport, ncol = n_Dport)
dimnames(missing) = list(top_Sport, top_Dport)
for (s in 1:n_Sport){
  for (d in 1:n_Dport){
    if(freqs[s,d] == 0){
      missing[s,d] = 1
    }
    else{
      missing[s,d] = 0
    }
  }
}
levelplot(missing,xlab = "Source Ports", ylab = "Destination Ports", scales=list(x=list(rot=90)),
          main = "Missingness in Ports Matrix", col.regions=gray(0:1,1),cuts=1)
```

The above matrix represents the missingness in the port combinations for pairings of the top 20 most used source ports and destination ports. The black cells represent missingness; of the 400 cells in the matrix, 295 (73.75%) of cells are missing observations. This single matrix slice can be extrapolated to missingness in ports throughout the entire tensor because the dataset is collected in a way such that either all four continuous features are observed, or none are observed. It is important to note that missingness is not uniform across source and destination port combinations. In the event that an entire row or column of port combinations is missing, the port at that respective index will need to be discarded from the simulation procedure because all three techniques depend on the row and column effects when simulating a value for a missing cell. 

## Port Connections

```{r, echo=FALSE, fig.height = 7, fig.width = 7}
### PORTS NETWORK GRAPH
# net = graph_from_adjacency_matrix(missing)
# net <- simplify(net, remove.multiple = T, remove.loops = T) 
# 
# 
# # Compute node degrees (#links) and use that to set node size:
# deg <- degree(net, mode="all")
# V(net)$size <- deg*3
# # We could also use the audience size value:
# V(net)$size <- V(net)$audience.size*0.6
# 
# E(net)$width <- E(net)$weight/6
# #change arrow size and edge color:
# E(net)$arrow.size <- .2
# E(net)$edge.color <- "gray80"
# plot(net, edge.arrow.size=.2, edge.color="orange",
#      vertex.color="orange", vertex.frame.color="#ffffff",
#      vertex.label.color="black") 


# Dataframe way
test = melt(means)
test = na.omit(test)
g = graph.data.frame(test)

V(g)$type <- 1
V(g)[name %in% top_Dport]$type <- 2
V(g)[name %in% top_Sport]$type <- 3
deg <- degree(g, mode="all")
V(g)$size <- deg
E(g)$width <- E(g)$weight * 3
E(g)$arrow.size <- .2
g = simplify(g, remove.multiple = T, remove.loops = T) 
shape <- c("circle", "square", "circle")
col <- c("orange", "steelblue","red")
plot(g,
     vertex.color = col[V(g)$type],
     vertex.shape = shape[V(g)$type],
     edge.arrow.size = 0.2
)
```

The above ports network graph displays the pairings between the top twenty source ports (red circles) and the top destination ports (blue squares). When a square and circle are connected it represents that there exist observations for this particular port combination in the dataset. The size of each node reflects the number of paired observations that were observed using that particular port. Clearly, not every source port is paired with a destination port and vice versa (not every node is connected to every other node). These missing combinations reflect missing cells in the $T$ tensor, and consequently they correspond to the combinations that require values to be simulated. 

## Varied Sample Sizes

```{r, echo = FALSE}
### MATRIX VISUALIZATIONS
# https://stackoverflow.com/questions/5453336/plot-correlation-matrix-into-a-graph
# levelplot(means, xlab = "Source Ports", ylab = "Destination Ports", 
#           main = "Means of SrcBytes By Ports", 
#           col.regions=rgb.palette(120))
freqs20 = readRDS("data/freqs_20.RDS")
levelplot(log(freqs20), xlab = "Source Ports", ylab = "Destination Ports",  
          col.regions = heat.colors(16)[length(heat.colors(16)):1],
          main = "Sample Sizes of Port Combinations (Log Scale)",scales=list(x=list(rot=90)))
```

To better visualize the sample sizes first displayed in the network graph, the above heat plot represents the sample sizes for each source-destination port combination on a log scale for clarity. It is again clear there are certain combinations that have many observations (range of 35000 on the original scale), while most of the observations are 0, indicating no observations were observed and the corresponding cell in $T$ is missing, or near 0, indicating few observations were observed. The large variation in sample sizes again suggests a simulation technique that accounts for sample size of a missing cell's related row and column cells is necessary.

## Row and Column Properties

```{r, echo = FALSE, fig.height=9, fig.width=6.5}
sbyte_means = readRDS("data/means_SB.rds")[1:20, 1:20]
spkt_means = readRDS("data/means_SP.rds")[1:20, 1:20]
dbyte_means = readRDS("data/means_DB.rds")[1:20, 1:20]
dpkt_means = readRDS("data/means_DP.rds")[1:20, 1:20]
### ROW MEANS/COL MEANS VISUALIZATION
sb_row_means = rowMeans(sbyte_means, na.rm = TRUE)
sb_col_means = colMeans(sbyte_means, na.rm = TRUE)

sp_row_means = rowMeans(spkt_means, na.rm = TRUE)
sp_col_means = colMeans(spkt_means, na.rm = TRUE)

db_row_means = rowMeans(dbyte_means, na.rm = TRUE)
db_col_means = colMeans(dbyte_means, na.rm = TRUE)

dp_row_means = rowMeans(dpkt_means, na.rm = TRUE)
dp_col_means = colMeans(dpkt_means, na.rm = TRUE)
#geombarplot, ggextra - grid arrange use ggplot use ggplot
par(mfrow=c(4,2))
barplot(sb_row_means, las=2, xlab = "Source Port", cex.names=0.8,
        main = "Row Means of SrcBytes Matrix", col = "green")
barplot(sb_col_means, las=2, xlab = "Destination Port", cex.names=0.8,
        main = "Column Means of SrcBytes Matrix", col = "orange")

barplot(sp_row_means, las=2, xlab = "Source Port", cex.names=0.8,
        main = "Row Means of SrcPkts Matrix", col = "green")
barplot(sp_col_means, las=2, xlab = "Destination Port",cex.names=0.8,
        main = "Column Means of SrcPkts Matrix", col = "orange")

barplot(db_row_means, las=2, xlab = "Source Port", cex.names=0.8,
        main = "Row Means of DstBytes Matrix", col = "green")
barplot(db_col_means, las=2, xlab = "Destination Port", cex.names=0.8,
        main = "Column Means of DstBytes Matrix", col = "orange")

barplot(dp_row_means, las=2, xlab = "Source Port", cex.names=0.8,
        main = "Row Means of DstPkts Matrix", col = "green")
barplot(dp_col_means, las=2, xlab = "Destination Port",cex.names=0.8,
        main = "Column Means of DstPkts Matrix", col = "orange")

```

The bar plots above represent the row and column means of the continuous features for each slice of the tensor. These row means and column means inform simulation techniques for the missing cells within those respective rows and columns. There exist clear outliers in the means for certain rows and columns. This outlier behavior is undoubtedly caused by outliers existing within the cells in that particular row or column. These outliers exist because each cell represents the mean of all observations that occured within a particular port combination, regardless of sample size (i.e. some cells may have a few large observations, resulting in a large mean that skews the cell's row and column mean). Thus, cells that only have a few observed observations have a disproportionately large effect on their respective row and column mean.

These outliers may cause problems with simulating missing values in that row or column because the outliers will have a disproportionately large effect on the simulated value than the other observations, which are more close to the median in the missing value's row or column.  This behavior suggests that the completion techniques that take into account variances among the row and column means and the number of samples observed for each port combination will result in more accurate estimations for the missing port combinations.

## Correlations
```{r, echo = FALSE, fig.height = 5, fig.width = 5}
### CORRELATIONS
kendall_cors = matrix(c(1.0000000, 0.7227563, 0.5729918, 0.6367907,
                0.7227563, 1.0000000, 0.7425211, 0.8108184,
                0.5729918, 0.7425211, 1.0000000, 0.8827988,
                0.6367907, 0.8108184, 0.8827988, 1.0000000),
                nrow = 4, ncol = 4)
rownames(kendall_cors) = c("SrcBytes", "SrcPkts", "DstBytes", "DstPkts")
colnames(kendall_cors) = c("SrcBytes", "SrcPkts", "DstBytes", "DstPkts")
levelplot(kendall_cors, main = "Kendall Correlations Between Continuous 
          Features", xlab = "", ylab = "")
#continuous_data = subset(argus, select = c("SrcBytes", "SrcPkts", "DstBytes", "DstPkts"))
#sub_data = continuous_data[1:50000,]
#cor(continuous_data, method = "kendall")

#           SrcBytes   SrcPkts  DstBytes   DstPkts
# SrcBytes 1.0000000 0.7227563 0.5729918 0.6367907
# SrcPkts  0.7227563 1.0000000 0.7425211 0.8108184
# DstBytes 0.5729918 0.7425211 1.0000000 0.8827988
# DstPkts  0.6367907 0.8108184 0.8827988 1.0000000

```

The matrix above describes the Kendall rank correlations (commonly referred to as Kendall's tau coefficent) between the four continuous features in the dataset. Intuitively, the Kendall correlation between two features will be high when observations have a similar rank (i.e. relative position label of observations within the variable: 1st, 2nd, 3rd, etc.) between the two variables, and low when observations have a dissimilar rank between the two variables. The range of correlations is [-1, 1]. Kendall correlation was selected as a measure because it evaluates ranks between observations, as opposed to Pearson, which is more susceptible to outliers in the dataset (large byte and packet observations in the continuous features skewed the Pearson measures). 

It is clear there exist strong correlations between the four continuous features, DstBytes and DstPkts in particular. This behavior suggests a technique that simulates estimates for the missing cells with all four features considered at once (i.e. a technique that simulates the entire tensor as a whole rather than in slices) will also be valuable.

Additional data analysis, including principal component analysis of the cells and the exploration of the different scale transformations, can be found in the Appendix.

<!-- ### Principal Component Analysis/Dimensionality Reduction -->

<!-- ```{r, echo=FALSE} -->
<!-- nscore = function(x) { -->
<!--    # Takes a vector of values x and calculates their normal scores. Returns  -->
<!--    # a list with the scores and an ordered table of original values and -->
<!--    # scores, which is useful as a back-transform table. See backtr(). -->
<!--    nscore = qqnorm(x, plot.it = FALSE)$x  # normal score  -->
<!--    trn.table = data.frame(x=sort(x),nscore=sort(nscore)) -->
<!--    return (list(nscore=nscore, trn.table=trn.table)) -->
<!-- } -->
<!-- #principal component analysis and visualizing results -->
<!-- pca_analysis = function(SrcBytes, SrcPkts, DstBytes, DstPkts){ -->
<!--   pca_cont_vars = cbind(SrcBytes, SrcPkts, DstBytes, DstPkts) -->
<!--   pca = prcomp(pca_cont_vars, center = TRUE, scale. = TRUE) -->
<!--   print(pca$rotation) -->
<!--   print((summary(pca))) -->
<!--   #screeplot(pca, type="lines",col=3) -->
<!--   g = ggbiplot(pca, obs.scale = 1, var.scale = 1, -->
<!--                 ellipse = TRUE, -->
<!--                 circle = TRUE) -->
<!--   g = g + scale_color_discrete(name = '') -->
<!--   g = g + theme(legend.direction = 'horizontal', -->
<!--                  legend.position = 'top') -->
<!--   return(g) -->
<!-- } -->
<!-- combinations = readRDS("data/pca_combo.rds") -->
<!-- combo_table = combinations[1] -->
<!-- combo_table = transform(combo_table, -->
<!--                         SrcBytes = as.numeric(SrcBytes), -->
<!--                         SrcPkts = as.numeric(SrcPkts), -->
<!--                         DstBytes = as.numeric(DstBytes), -->
<!--                         DstPkts = as.numeric(DstPkts)) -->
<!-- # cat("Sport:", combo_table$Sport[1],"\t") -->
<!-- # cat("Dport:", combo_table$Dport[1],"\n") -->
<!-- SrcBytes_norm =  nscore(combo_table$SrcBytes)$nscore -->
<!-- SrcPkts_norm =  nscore(combo_table$SrcPkts)$nscore -->
<!-- DstBytes_norm =  nscore(combo_table$DstBytes)$nscore -->
<!-- DstPkts_norm =  nscore(combo_table$DstPkts)$nscore -->
<!-- g1 = pca_analysis(SrcBytes_norm, SrcPkts_norm, DstBytes_norm, DstPkts_norm) -->

<!-- combo_table = combinations[20] -->
<!-- combo_table = transform(combo_table, -->
<!--                         SrcBytes = as.numeric(SrcBytes), -->
<!--                         SrcPkts = as.numeric(SrcPkts), -->
<!--                         DstBytes = as.numeric(DstBytes), -->
<!--                         DstPkts = as.numeric(DstPkts)) -->
<!-- # cat("Sport:", combo_table$Sport[1],"\t") -->
<!-- # cat("Dport:", combo_table$Dport[1],"\n") -->
<!-- SrcBytes_norm =  nscore(combo_table$SrcBytes)$nscore -->
<!-- SrcPkts_norm =  nscore(combo_table$SrcPkts)$nscore -->
<!-- DstBytes_norm =  nscore(combo_table$DstBytes)$nscore -->
<!-- DstPkts_norm =  nscore(combo_table$DstPkts)$nscore -->
<!-- g20 = pca_analysis(SrcBytes_norm, SrcPkts_norm, DstBytes_norm, DstPkts_norm) -->

<!-- grid.arrange(g1, g20, ncol = 2) -->

<!-- ``` -->



